\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-01-28 analysis/manuscript.Rmd
Writing methods

---
title: "Overview"
output: github_document
always_allow_html: true
date: "`r Sys.Date()`"
---

@@

@@

```{r}
publisher_league_tbl |>
  mutate(across(contains(c("perc", "cum")), ~ paste(round(., 3) * 100))) |>
  kbl(
    caption ="Hybrid open access through transformative agreements market shares 2018-2022",
    col.names = c(
      "Publisher",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%"
    ),
    format.args = list(big.mark = ',')
  ) %>%
  kable_classic() %>%
  add_header_above(c(
    " " = 1,
    "Hybrid journals" = 3,
    "Articles" = 3,
    "OA articles" = 3,
    "TA OA articles" = 3
  ))
```

@@

subject_tbl |>
  mutate(across(contains(c("perc", "cum")), ~ paste(round(., 3) * 100))) |>
  kbl(
    caption ="Hybrid open access through transformative agreements by journal subject 2018-2022",
    col.names = c(
      "Journal subject",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%"
    ),
    format.args = list(big.mark = ',')
  ) %>%
  kable_classic() %>%
  add_header_above(c(
    " " = 1,
    "Hybrid journals" = 3,
    "Articles" = 3,
    "OA articles" = 3,
    "TA OA articles" = 3
  ))
```

>>>>

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: apa6.sty
csl: apa.csl
bibliography: references.bib
---

@@

```
# Introduction

# Methods and data

This study combines data from multiple publicly available data sources as visualised in Figure 1. 
Data gathering involved transformative agreement data retrieved from the cOAlition S Journal Checker Tool, which provided information about journal portfolios and participating institutions. 
After identification of hybrid journals, Crossref served as the primary data source for article-level metadata including Creative Commons (CC) license information to determine open access articles on publisher websites.
OpenAlex provided first author affiliation metadata that were subsequently linked to the transformative agreement data to identify open access articles published through these agreements. 
In total, the compiled data set consists of `r hoaddata::jn_ind |> filter(!cr_year %in% c("2017", "2023")) |> distinct(issn_l, cr_year, jn_all) |> pull(jn_all) |> sum()` articles published in `r hoaddata::jn_ind |> filter(!cr_year %in% c("2017", "2023")) |> distinct(issn_l) |> nrow()` hybrid journals between 2018 and 2022.

```{r data_workflow, echo=FALSE, fig.cap="Data collection workflow to obtain article-level transformative agreement data"}
knitr::include_graphics(here::here("analysis", "data_collection_workflow.pdf"))
```

## Transformative agreement and hybrid journal data

```{r journal_oa_exclusion}
oa_journal_status <- readr::read_csv(here::here("data", "oa_journal_status.csv"))

# OA journal total
n_oa_journals <- oa_journal_status %>%
  filter(openalex == TRUE | doaj == TRUE | bielefeld == TRUE) %>% 
  distinct(issn_l) %>%
  nrow()
# By source
#
# GOLD OA
bielefeld <- oa_journal_status %>%
  filter(bielefeld == TRUE) %>% 
  distinct(issn_l)  %>%
  pull()
# DOAJ
doaj <- oa_journal_status %>%
  filter(doaj == TRUE) %>% 
  distinct(issn_l) %>%
  pull()
# OpenAlex venue
openalex <- oa_journal_status %>%
  filter(openalex == TRUE) %>% 
  distinct(issn_l) %>%
  pull()
my_venn <- list(`GOLD OA` = bielefeld, `DOAJ` = doaj, `OpenAlex` = openalex)

p <-
  ggvenn::ggvenn(
    my_venn,
    fill_color = c("grey55",
          "steelblue2",
          "lightsalmon"),
    digits = 0,
    text_color = "black",
    text_size = 3,
    set_name_size = 4
  )
```

Data gathering started with obtaining journals included in transformative agreements using the cOAlition S’ Journal Checker Tool publicly available Transformative Agreement Data dump. 
The dump consisted of multiple online Google spreadsheets where each data file represented one agreement listed in the ESAC Transformative Agreement Registry. 
Each file comprised information about journals and institutions involved per agreement. 
Because the Journal Checker Tool only collected data about active transformative agreements, four different snapshots were used for this analysis:the archived version from July 2021, July 2022, May 2023 and the most current dump downloaded on 11 December 2022. 
This ensured that transformative agreements, which ended in 2021, were also included in the analysis. 
Agreements ending before 2021 could not be obtained.
Overall, the combined Transformative Agreement Data dumps contained `r  length(unique(hoaddata::jct_hybrid_jns$esac_id))` agreements.

The Transformative Agreement Data dumps linked agreements to journals represented by journal names and ISSNs. 
After mapping ISSN variants to the corresponding linking ISSN (ISSN-L) as provided by The ISSN International Centre.
Journals were associated to publishers using the ESAC ID, a unique identifier for transformative agreements in the ESAC Transformative Agreement Registry.

Because transformative agreements can include both full open access and hybrid journals, journal data were complemented with information about a journal's open access status using multiple sources: the Directory of Open Access Journals (DOAJ) downloaded on 4 October 2022, OpenAlex (September 2022 dump) and the the Bielefeld list of GOLD OA journals. 
In total, `r n_oa_journals` full open access journals were excluded.
Figure shows a substantial overlap between the three data sources of
`r round(length(unique(Reduce(intersect, my_venn))) / length(unique(Reduce(union, my_venn)))* 100,0)`%.
However, combining different data sources considerably extended the journal matching to exclude full open access journals from transformative agreements.
The Gold OA journals dataset alone adds `r length(unique(Reduce(setdiff, my_venn)))` journals, while the DOAJ comprised `r length(setdiff(my_venn$DOAJ, union(my_venn[["GOLD OA"]], my_venn$OpenAlex)))` full open access journals not listed in either of the other two sources. 
These full open access journals were mostly launched in 2022.

## Article metadata

## Data Analysis

@@

```{r}
publisher_league_tbl |>
  mutate(across(contains(c("perc", "cum")), ~ paste(round(., 3) * 100))) |>
  kbl(
    booktabs = TRUE, format.args = list(big.mark = ","), digits = 1, escape = TRUE, position = "H",
    caption ="Hybrid open access through transformative agreements market shares 2018-2022",
    col.names = c(
      "Publisher",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%"
    )
  ) |>
  kable_styling(font_size = 10, position = "center") |>
  add_header_above(c(
    " " = 1,
    "Hybrid journals" = 3,
    "Articles" = 3,
    "OA articles" = 3,
    "TA OA articles" = 3
  ))
```

@@

subject_tbl |>
  mutate(across(contains(c("perc", "cum")), ~ paste(round(., 3) * 100))) |>
  kbl(
    booktabs = TRUE, format.args = list(big.mark = ","), digits = 1, escape = TRUE, position = "H",
    caption ="Hybrid open access through transformative agreements by journal subject 2018-2022",
    col.names = c(
      "Journal subject",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%",
      "Total",
      "%",
      "C%"
    )
  ) |>
  add_header_above(c(
    " " = 1,
    "Hybrid journals" = 3,
    "Articles" = 3,
    "OA articles" = 3,
    "TA OA articles" = 3
  ))
```
